Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-22 manuscript/HIAs.qmd
vectorize infrastructure index

myCRS <- 25832
```

@@

```{r mireDataTestCrop}
file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
# This is a big TIF!

mire_test <- read_stars(file, proxy = T) |>
  setNames("mire") |>
  st_downsample(100) |>
  mutate(mire = na_if(mire, 0)) |>
  st_as_stars()

@@

# The cropping is correct
tmap_mode("plot")
tm_shape(mire_test)+
  tm_raster(palette = "red")+
  tm_shape(outline)+
  tm_borders()+
  tm_shape(mire_test_nf)+
  tm_raster(col="red")
```

@@

mire_nf <- read_stars(file, proxy = T) |>
  setNames("mire") |>
  #mutate(mire = na_if(mire, 0)) |>
  st_crop(nf) |>
  st_as_stars()

@@

# crop and mask (very fast!)
mire_nf_terra <- terra::rast(file) |>
  crop(nf_vect) |>
  terra::mask(nf_vect)

# Plot to check overlap
ggplot()+
  geom_spatraster(data = mire_nf_terra)+
  geom_spatvector(data = nf_vect,
                  fill = NA)
# The cropping and masking worked. 
#tmap_mode("plot")
#tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_nf_terra)+
#  tm_raster(col="red")

# I ilke the stars, sf and tmap combo better, so I return to stars
mire_nf_stars <- mire_nf_terra |>
  st_as_stars()

tm_shape(mire_test)+
  tm_raster(palette = "red")+
  tm_shape(outline)+
  tm_borders()+
  tm_shape(mire_nf_stars)+
  tm_raster(col="red")
# OK

@@

@@

@@

We chose three municipalities in Norway to test out the indicator @fig-studyLocation.
These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).

```{r}
# calculate terrestrial area
terrestrial <- terrestrial |> 
  mutate(t_area_km = 
           round(units::drop_units(area_t * 1e-6))) |>
  as_tibble()

# calculate area of survey coverage maps
dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_cast("POLYGON") |>
  group_by(Municipality) |>
  summarise(geometry = st_union(geometry)) |>
  mutate(dk_area_km = geometry |> st_area(),
         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))

# number and area of mire polygons
nature3_tbl <- nature3 |>
  group_by(Municipality) |>
  summarise(n = n()) |>
  as_tibble()

saveRDS(muni_tbl, "../data/municipality_stats.RDS")

@@

| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |
| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |
| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |

>>>>

myCRS <- 25832

```{r paths}
# Conditional file directory
dir <- substr(getwd(), 1,2)

mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"

path_infrastructure <- ifelse(dir == "C:", 
        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")
```

@@

```{r mireDataTestCrop}

# TEsting the cropping proceedure using a downsampled starts object 
mire_test <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  st_downsample(100) |>
  mutate(mire = na_if(mire, 0)) |>
  st_as_stars()

@@

# The cropping is correct:
#tmap_mode("plot")
#tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_test_nf)+
#  tm_raster(col="red")
```

@@

mire_nf <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  #mutate(mire = na_if(mire, 0)) |>
  st_crop(nf) |>
  st_as_stars()

@@

# crop and mask (very fast!)
mire_nf_terra <- terra::rast(mireFile) |>
  terra::crop(nf_vect) |>
  terra::mask(nf_vect)

# Plot to check overlap
#ggplot()+
#  geom_spatraster(data = mire_nf_terra)+
#  geom_spatvector(data = nf_vect,
#                  fill = NA)
# The cropping and masking worked.

# I like the stars, sf and tmap combo better, so I return to stars
mire_nf_stars <- mire_nf_terra |>
  st_as_stars()

#tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_nf_stars)+
#  tm_raster(col="red")
# OK

@@

```{r dk2}
# calculate terrestrial area
terrestrial <- terrestrial |> 
  mutate(t_area_km = 
           round(units::drop_units(area_t * 1e-6))) |>
  as_tibble()

# calculate area of survey coverage maps
dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_cast("POLYGON") |>
  group_by(Municipality) |>
  summarise(geometry = st_union(geometry)) |>
  mutate(dk_area_km = geometry |> st_area(),
         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))
```

@@

```{r infrastructureIndex}
# this data is on a 100x100m grid
infra <- stars::read_stars(path_infrastructure) |>
  # , which is mor then we need - warp it to 1x1km
  st_warp(cellsize = c(1000, 1000), 
                                   crs = st_crs(nf), 
                                   use_gdal = TRUE,
                                   method = "average") |>
  setNames("infrastructureIndex") |>
  st_transform(myCRS) |>
   mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 12 ~ 2,
    infrastructureIndex >= 12 ~ 3)) |>
  st_crop(outline)

# This step might seem rather stupid. We want to vectorize a rather large raster. This makes it a quite big data object. The reason is that there is no really good way to burn polygon data on to raster grid cells after the disuse of the raster package. It was not straight forward then either. But calculating intersections between polygons is very fast and easy.

infra <-  eaTools::ea_homogeneous_area(infra, 
                                groups = infrastructureIndex)

path <- ifelse(dir == "C:", 
        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/")

saveRDS(infra, paste0(path, "infrastructureIndex_discrete_vectorized.rds"))
```

```{r muni-tbl}
# number and area of mire polygons
nature3_tbl <- nature3 |>
  group_by(Municipality) |>
  summarise(n = n()) |>
  as_tibble()

saveRDS(muni_tbl, "../data/municipality_stats.RDS")
```

@@

We chose three municipalities in Norway to test out the indicators (@fig-studyLocation).
These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).

@@

| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value  |
|-------|----------|----------|----------|----------|----------|----------|----------|
| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |                                  |
| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |                                  |
| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |                                  |
